Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2021-01-15 Nicolas.tex
MR relecture and MB corrections. Bold to add.

@@

The $H_0$ tension has received a lot of attention, as it could be a sign of new
fundamental physics. Yet, no simple solution is able to accommodate this $H_0$
tension when accounting for all other probes \citep{knox2019}, but see e.g.
\cite{poulin2019}. Alternatively, systematic effects affecting one or several
of the aforementioned analyses might explain at least some of this tension.
\cite{rigault2015} suggested that SNe Ia from the Cepheid calibrator sample
differ by construction from those in the Hubble flow sample, as the former
strongly favors young stellar populations, while the latter does not. This
selection effect would impact the derivation of $H_0$ if SNe~Ia from young and
older environments differ in average standardized magnitudes.

%For the last decades, numerous analyses have studied 
The relation between SNe~Ia
and their host galaxy properties has been studied extensively. The first key finding
was that that the standardized SNe~Ia magnitudes
significantly depend on the host galaxy stellar mass, with SNe~Ia from high-mass host 
galaxies being
brighter on average \cite[e.g.][]{kelly2010, sullivan2010, childress2013,
betoule2014, rigault2020, kim19}. This mass-step correction is currently used in
cosmological analyses \citep[e.g.][]{betoule2014, scolnic2018a}, including for
deriving $H_0$ \citep{riess2016, riess2019}. Yet, the underlying connection
between the SNe and their host galaxies remains unclear, particularly when using global properties such
as the host stellar mass. This in turn raises the question of the accuracy of such
corrections since host galaxy global properties evolve with redshift. 
More recently, studies have used the local SN host galaxy environment to probe
more direct connections between the SNe and their host galaxy environments \citep{rigault2013},
showing that local age tracers such as the Local specific Star Formation Rate
(LsSFR) or the local color are more strongly correlated with the standardized SN
magnitude \citep{rigault2020, roman2018, kim18}. Further reinforcing this
connection is the identification of SN~Ia spectral features that are 
correlated with LsSFR \citep{nordin2018}. These results suggest age as the driving
parameter underlying the mass-step. If true, this would have a significant
impact for cosmology, since the environmental correction to apply to SN
standardization could strongly vary with redshift. \citep{rigault2013,
childress2014, scolnic2018a}. Yet, the importance of local SN environmental
studies remains highly debated \cite[e.g.][]{jones2015, jones2019} and
especially the impact of such an astrophysical bias has on the derivation of
$H_0$ \citep{jones2015, riess2016, riess2018, rose2019}.

The concept of the SN~Ia age dichotomy arose with the study of the SN~Ia rate.
\cite{mannucci2005, scannapieco2005, sullivan2006, aubourg2008} have shown that
the relative SNe~Ia rate in galaxies could be explained if two populations
existed, one young, following the host star formation activity, and one old
following the host stellar mass (the so called ``prompt and delayed'' or ``A+B''
model). In \cite{rigault2020} we used the LsSFR to classify which are the
younger (those with a high LsSFR) and which are the older (those with low
LsSFR). Furthermore, since the first SNe~Ia host analyses, the SN stretch has
been known to be strongly correlated with the SN host galaxy properties
\citep{hamuy1996, hamuy2000}. This correlation that has been extensively confirmed
since \citep[e.g.][]{neill2009, sullivan2010, lampeitl2010, kelly2010,
gupta2011, dandrea2011, childress2013, rigault2013, pan2014, kim19}. Following
the ``A+B'' model and the connection between SN stretch and host galaxy properties,
\cite{howell2007} first discussed the potential redshift drift of the SN stretch
distribution. In this paper we revisit this question using the most recent SNe~Ia
datasets.

In this paper, we take a step aside to probe the validity of our modeling of the
SN population, which we claim to be constituted of two age-populations
\citep{rigault2013,rigault2015,rigault2020}: one old and one younger, the former
having on average lower lightcurve stretches and being brighter after
standardization. We use the correlation between the SN age, as probed by the
LsSFR, and the SN stretch to model the expected evolution of the underlying SN
stretch distribution as a function of redshift. This modeling relies on three
assumptions: (1) there are two distinct populations of SNe~Ia; (2) the relative
fraction of each of these populations as a function of redshift follows the
model presented in \cite{rigault2020} and (3) the underlying distribution of
stretch for each age sample is constant. This paper aims at testing this
specific model with datasets from the literature. 
% Note that the progenitor age
% as traced by the LsSFR seems to capture physical feature intrinsic to the
%progenitor and/or explosion mechanism that the stretch alone is not capturing
% \citep{nordin2018}.

We present in Section~\ref{sec:sample} the sample we are using for this
analysis, derived from the Pantheon catalog \citep{scolnic2018a}. We discuss the
importance of obtaining a ``complete'' sample, i.e.\ representative of the true
underlying SNe Ia distribution, and how we build one from the Pantheon sample.
We then present in Section~\ref{sec:modeling} our modeling of the distribution
of stretch and our results are presented in Section~\ref{sec:results}. In this
section we test whether the SN stretch distribution evolves as a function of
redshift and if the aforementioned age model is in good agreement with this
evolution. We briefly discuss these results in the context of SN cosmology in
Section~\ref{sec:discussion} and we conclude in Section~\ref{sec:ccl}.

@@

Assuming sufficient (and unbiased) spectroscopic follow-up for acquiring type SN types
and host galaxy redshifts, the observation selection effects of magnitude-limited surveys should be
negligible below a given redshift at which even the faintest normal SNe~Ia can be
observed. \textbf{Aiding in the construction of nearly volume-limited 
subsamples is the fact that the SN~Ia population trails off towards fainter SNe~Ia.
A complication is that complete spectroscopic follow-up has not always been
the norm, as discussed below.}
In contrast, targeted surveys have highly complex observational
selection functions and so are discarded from our analysis. High-redshift SN cosmology
samples, such as Pantheon, are predominately from magnitude-limited surveys from
which volume-limited SN~Ia subsamples can be constructed.
%\textbf{which have easier to model observational selection functions}.

@@

\textbf{We will thus consider a set of cuts that will define a first fiducial
    sample, taking the limits as initially suggested by the previous
    completeness analysis. \textbf{However,} as this solution might be an overly
    simplified way to create a complete sample, e.g., because it ignores spectroscopic follow-up
    in efficiency for redshifts below $z_{lim}$, we also consider another set of cuts to define a so-called
    ``conservative'' sample. The latter is smaller and therefore less statistically
    constraining, but also even less prone to observational selection effects. If the
    redshift drift is still significant in the conservative sample, it would be
    even more meaningful in a carefully-tailored selection-free sample. These
    samples are adequate for the goal
    of this study, which is to develop a first implementation of a model for drift in
    SN Ia properties. If fruitful, the sample selection can later be refined, if needed, with a
    more detailed
    %drift modeling before refining it on a more complex, yet more accurate,
%follow-up analysis also accounting for precise 
model of the observational selection, e.g., using
the SNANA package \citep{SNANA}.}

@@

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift
distribution of these three surveys is shown in Fig.~\ref{fig:cuts}. As
expected, the selected redshift limits \textbf{are roughly located slightly
before the peak of these histograms. In Section 2.2 we validate that these
redshift limits are effective for constructing nearly volume-limited subsamples
from samples that were initially more closely magnitude-limited in their search
or spectroscopic follow-up.}

@@

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}. \textbf{We observe here
that the fraction of low-stretch SNe (typically $x_1 < -1$) appears to decrease as
a function of redshift; this is confirmed in Fig.~\ref{fig:modelall} in which}
the evolution of the mean stretch is shown, \textbf{with} the data split in
redshift bins of regular sample size. We see that SNe~Ia at higher redshift have
on average larger stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower
redshift ($-0.17\pm 0.10$ at $z\sim0.05$), suggesting that the underlying
stretch distribution is \textbf{evolving with redshift}.

@@

The stretch probability distribution function (pdf) of a given SN will be the
linear combination of the stretch distributions of these two population weighted
by its probability $y^i$ to be young (see Section~\ref{sec:basemodelapplied}).
But generally, the fraction of young SNe~Ia as a function of redshift is given
by $\delta(z)$ (see Eq.~\ref{eq:delta}) and therefore, our redshift drift model
of the underlying distribution of SNe~Ia as a function of redshift $X_1(z)$ is
given by:
\begin{align}\label{eq:stretchz}
    X_1(z) = \delta(z)&\times \mathcal{N}(\mu_1,\sigma_1{}^2) + \nonumber \\
    (1-\delta(z))&\times \left[ a\times\mathcal{N}(\mu_1,\sigma_1{}^2) +
    (1-a)\times\mathcal{N}(\mu_2,\sigma_2{}^2) \right]
\end{align}
\textbf{This is our Base drifting model.}

@@

When lacking direct LsSFR measurements (i.e. $p_y^i$), we can extend the
analysis to non-SNfactory samples by using the redshift-evolution of the
fraction $\delta(z)$ of young SNe~Ia (Eq.~\ref{eq:delta}) as a proxy for the
probability of a SN to be young. This still corresponds to minimizing
Eq.~\ref{eq:likelihood} with respect to the parameters
$\vec{\theta}\equiv(\mu_1, \mu_2, \sigma_1, \sigma_2, a)$ of the stretch
distribution $X_1$ (Eq.~\ref{eq:stretchz}), but this time assuming $y^i =
\delta(z^i)$ for any given SN~$i$.

For the rest of the analysis, we will therefore minimize Eq.~\ref{eq:likelihood}
using $p_y^i$ --- the probability for the SN $i$ to be young --- when available
(i.e. for SNfactory dataset), and $\delta(z^i)$ --- the expected fraction of
young SNe~Ia at the SN redshift $z^i$ --- otherwise.

@@

\bibitem[Poulin et al.(2019)]{poulin2019} Poulin, V., Smith, T.~L., Karwal, T.,
et al.\ 2019, \prl, 122, 221301

@@

\bibitem[{Rigault {et~al.}(2013)}]{rigault2013} Rigault, M., Copin, Y.,
Aldering, G., {et~al.} 2013, \aap, 560, A66

>>>>

\titlerunning{Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution}
\authorrunning{N.~Nicolas et al.}

@@

The $H_0$ tension has received a lot of attention, as it could be a sign of new
fundamental physics. Yet, no simple solution is able to accommodate this $H_0$
tension when accounting for all other probes \citep{knox2019}. Alternatively, systematic effects affecting one or several
of the aforementioned analyses might explain at least some of this tension.
\cite{rigault2015} suggested that SNe Ia from the Cepheid calibrator sample
differ by construction from those in the Hubble flow sample, as the former
strongly favors young stellar populations, while the latter does not. This
selection effect would impact the derivation of $H_0$ if SNe~Ia from young and
older environments differ in average standardized magnitudes.

%For the last decades, numerous analyses have studied 
The relation between SNe~Ia
and their host galaxy properties has been studied extensively. The first key finding
was that the standardized SNe~Ia magnitudes
significantly depend on the host galaxy stellar mass, with SNe~Ia from high-mass host 
galaxies being
brighter on average \cite[e.g.][]{kelly2010, sullivan2010, childress2013,
betoule2014, rigault2020, kim19}. This mass-step correction is currently used in
cosmological analyses \citep[e.g.][]{betoule2014, scolnic2018a}, including for
deriving $H_0$ \citep{riess2016, riess2019}. Yet, the underlying connection
between the SNe and their host galaxies remains unclear, particularly when using global properties such
as the host stellar mass. This in turn raises the question of the accuracy of such
corrections since host galaxy global properties evolve with redshift. 
More recently, studies have used the local SN host galaxy environment to probe
more direct connections between the SNe and their host galaxy environment \citep{rigault2013},
showing that local age tracers such as the Local specific Star Formation Rate
(LsSFR) or the local color are more strongly correlated with the standardized SN
magnitude \citep{roman2018, kim18,rigault2020}. Further reinforcing this
connection is the identification of SN~Ia spectral features that are 
correlated with LsSFR \citep{nordin2018}. These results suggest age as the driving
parameter underlying the mass-step. If true, this would have a significant
impact for cosmology, since the environmental correction to apply to SN
standardization could strongly vary with redshift. \citep{rigault2013,
childress2014, scolnic2018a}. Yet, the importance of local SN environmental
studies remains highly debated \cite[e.g.][]{jones2015, jones2019} and
especially the impact of such an astrophysical bias has on the derivation of
$H_0$ \citep{jones2015, riess2016, riess2018, rose2019}.

The concept of the SN~Ia age dichotomy arose with the study of the SN~Ia rate.
\cite{mannucci2005, scannapieco2005, sullivan2006, aubourg2008} have shown that
the relative SNe~Ia rate in galaxies could be explained if two populations
existed, one young, following the host star formation activity, and one old
following the host stellar mass (the so called ``prompt and delayed'' or ``A+B''
model). \cite{rigault2020} used the LsSFR to classify which are the
younger (those with a high LsSFR) and which are the older (those with low
LsSFR). Furthermore, since the first SNe~Ia host analyses, the SN stretch has
been known to be strongly correlated with the SN host galaxy properties
\citep{hamuy1996, hamuy2000}. This correlation that has been extensively confirmed
since \citep[e.g.][]{neill2009, sullivan2010, lampeitl2010, kelly2010,
gupta2011, dandrea2011, childress2013, rigault2013, pan2014, kim19}. Following
the ``A+B'' model and the connection between SN stretch and host galaxy properties,
\cite{howell2007} first discussed the potential redshift drift of the SN stretch
distribution. In this paper we revisit this question using the most recent SNe~Ia
datasets.

In this paper, we take a step aside \mr{from the cosmological analyses} to probe the validity of our modeling of the
SN population, which we claim to be constituted of two age-populations
\citep{rigault2013,rigault2015,rigault2020}: one old and one younger, the former
having on average lower lightcurve stretches and being brighter after
standardization. We use the correlation between the SN age, as probed by the
LsSFR, and the SN stretch to model the expected evolution of the underlying SN
stretch distribution as a function of redshift. This modeling relies on three
assumptions: (1) there are two distinct populations of SNe~Ia; (2) the relative
fraction of each of these populations as a function of redshift follows the
model presented in \cite{rigault2020} and (3) the underlying distribution of
stretch for each age sample is constant. This paper aims at testing this
specific model with datasets from the literature. 
% Note that the progenitor age
% as traced by the LsSFR seems to capture physical feature intrinsic to the
%progenitor and/or explosion mechanism that the stretch alone is not capturing
% \citep{nordin2018}.

We present in Section~\ref{sec:sample} the sample we are using for this
analysis, derived from the Pantheon catalog \citep{scolnic2018a}. We discuss the
importance of obtaining a ``complete'' sample, i.e.\ representative of the true
underlying SNe Ia distribution, and how we build one from the Pantheon sample.
We then present in Section~\ref{sec:modeling} our modeling of the distribution
of stretch and our results are presented in Section~\ref{sec:results}. In this
section, we test whether the SN stretch distribution evolves as a function of
redshift and if the aforementioned age model is in good agreement with this
evolution. We briefly discuss these results in the context of SN cosmology in
Section~\ref{sec:discussion} and we conclude in Section~\ref{sec:ccl}.

@@

Assuming sufficient (and unbiased) spectroscopic follow-up for acquiring SN types
and host galaxy redshifts, the observation selection effects of magnitude-limited surveys should be
negligible below a given redshift at which even the faintest normal SNe~Ia can be
observed. \textbf{Aiding in the construction of nearly volume-limited 
subsamples is the fact that the SN~Ia population trails off towards fainter SNe~Ia.
A complication is that complete spectroscopic follow-up has not always been
the norm, as discussed below.}
In contrast, targeted surveys have highly complex observational
selection functions and so are discarded from our analysis. High-redshift SN cosmology
samples, such as Pantheon, are predominately from magnitude-limited surveys from
which volume-limited SN~Ia subsamples can be constructed.
%\textbf{which have easier to model observational selection functions}.

@@

\textbf{We will thus consider a set of cuts that will define a first fiducial
    sample, taking the limits as initially suggested by the previous
    completeness analysis. \textbf{However,} as this solution might be an overly
    simplified way to create a complete sample (e.g., because it ignores spectroscopic follow-up
    in efficiency for redshifts below $z_{lim}$), we also consider another set of cuts to define a so-called
    ``conservative'' sample. The latter is smaller and therefore less statistically
    constraining, but also even less prone to observational selection effects. If the
    redshift drift is still significant in the conservative sample, it would be
    even more meaningful in a carefully-tailored selection-free sample. These
    samples are adequate for the goal
    of this study, which is to develop a first implementation of a model for drift in
    SN Ia properties. If fruitful, the sample selection can later be refined, if needed, with a
    more detailed
    %drift modeling before refining it on a more complex, yet more accurate,
%follow-up analysis also accounting for precise 
model of the observational selection, e.g., using
the SNANA package \citep{SNANA}.}

@@

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift
distribution of these three surveys is shown in Fig.~\ref{fig:cuts}. As
expected, the selected redshift limits \textbf{are roughly located slightly
before the peak of these histograms. In Section \ref{ssec:verify} we validate that these
redshift limits are effective for constructing nearly volume-limited subsamples
from samples that were initially more closely magnitude-limited in their search
or spectroscopic follow-up.}

@@

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}. \textbf{We observe here
that the fraction of low-stretch SNe (typically $x_1 < -1$) appears to decrease as
a function of redshift; this is confirmed in Fig.~\ref{fig:modelall}, in which}
the evolution of the mean stretch is shown, \textbf{with} the data split in
redshift bins of regular sample size. We see that SNe~Ia at higher redshift have
on average larger stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower
redshift ($-0.17\pm 0.10$ at $z\sim0.05$), suggesting that the underlying
stretch distribution is \textbf{evolving with redshift}.

@@

The stretch probability distribution function (pdf) of a given SN will be the
linear combination of the stretch distributions of these two population weighted
by its probability $y^i$ to be young (see Section~\ref{sec:basemodelapplied}).
But generally, the fraction of young SNe~Ia as a function of redshift is given
by $\delta(z)$ (see Eq.~\ref{eq:delta}) and therefore, our redshift drift model
of the underlying \textbf{stretch} distribution of SNe~Ia as a function of redshift $X_1(z)$ is
given by:
\begin{align}\label{eq:stretchz}
    X_1(z) = \delta(z)&\times \mathcal{N}(\mu_1,\sigma_1{}^2) + \nonumber \\
    (1-\delta(z))&\times \left[ a\times\mathcal{N}(\mu_1,\sigma_1{}^2) +
    (1-a)\times\mathcal{N}(\mu_2,\sigma_2{}^2) \right]
\end{align}
\textbf{This is our Base drifting model.}

@@

When lacking direct LsSFR measurements (i.e. $p_y^i$), we can extend the
analysis to non-SNfactory samples by using the redshift-evolution of the
fraction $\delta(z)$ of young SNe~Ia (Eq.~\ref{eq:delta}) as a proxy for the
probability of a SN to be young. This still corresponds to minimizing
Eq.~\ref{eq:likelihood} with respect to the parameters
$\vec{\theta}\equiv(\mu_1, \mu_2, \sigma_1, \sigma_2, a)$ of the stretch
distribution $X_1$ (Eq.~\ref{eq:stretchz}), but this time assuming $y^i =
\delta(z^i)$ for any given SN~$i$. 
\\
\\
For the rest of the analysis, we will therefore minimize Eq.~\ref{eq:likelihood}
using $p_y^i$ --- the probability for the SN $i$ to be young --- when available
(i.e. for SNfactory dataset), and $\delta(z^i)$ --- the expected fraction of
young SNe~Ia at the SN redshift $z^i$ --- otherwise.

@@

%\bibitem[Poulin et al.(2019)]{poulin2019} Poulin, V., Smith, T.~L., Karwal, T., et al.\ 2019, \prl, 122, 221301

@@

\bibitem[{Rigault et al.(2013)}]{rigault2013} Rigault, M., Copin, Y.,
Aldering, G., {et~al.} 2013, \aap, 560, A66
